---
title: 'phinterval: An R package for representing and manipulating timespans with gaps'
tags:
  - R
  - dates
  - timespans
  - set operations
authors:
  - name: Ethan Sansom
    orcid: 0009-0000-1573-0186
    affiliation: 1
affiliations:
  - name: "Department of Statistical Sciences, University of Toronto"
    index: 1
date: "2025-11-12"
bibliography: paper.bib
output:
  #rticles::joss_article:
  #  keep_tex: false
  md_document:
     preserve_yaml: TRUE
     variant: "markdown_strict"
journal: JOSS
---

# Summary

`phinterval` is an R [@R] package for representing and manipulating time spans that may contain gaps. It implements the *phinterval* vector class, designed as an extension of the `lubridate` [@lubridate] package's *Interval* class, to represent continuous, disjoint, empty, and unknown spans of time.

Functionality for manipulating these spans includes:

-   Performing set operations: union, intersection, difference, and complement.
-   Merging overlapping or adjacent intervals into non-overlapping sets of time spans.
-   Testing whether time spans, dates, or times fall within one another or overlap.

# Statement of Need

Because of the complexities of accurately representing dates and times, including adjustments for time zones, daylight saving transitions, and leap years or seconds, manipulating time spans is a common source of frustration and error-prone code for analysts [@lubridate; @itsabouttime]. Several R packages, notably `lubridate` and `ivs` [@ivs], provide intuitive interfaces for representing and manipulating time spans that handle these complexities internally, reducing the cognitive load required of users and the likelihood of mistakes.

To the author's knowledge, however, no existing package supports empty or discontinuous time spans. Users encountering these spans - for example, the intersection of two non-overlapping intervals - receive an error (as in `ivs`) or unintuitive results (as in `lubridate`), forcing workarounds or potentially leading to uncaught mistakes. The `phinterval` package addresses this gap by providing explicit representations of disjoint and empty time spans, enabling operations such as union, intersection, and set difference to be performed accurately on arbitrary intervals. Its interface closely mirrors that of `lubridate`, and all `phinterval` functions accept `lubridate` *Interval* vectors as inputs, allowing analysts to safely integrate `phinterval` into their existing workflows and work with a broader range of temporal data.

# Examples

```{r dependencies, warning = FALSE, message = FALSE, echo = FALSE}
# pak::pkg_install("EthanSansom/phinterval")
library(phinterval)
library(lubridate)
```

To demonstrate the utility of the *phinterval* class, consider a modified example from the `lubridate` package vignettes. Suppose two colleagues are each taking a 5 day vacation: one to Greece in early January and the other to Brazil in mid-February.

```{r intervals}
greece <- interval(ymd("2020-01-01"), ymd("2020-01-06"))
brazil <- interval(ymd("2020-02-11"), ymd("2020-02-16"))
```

If we compute the union of these intervals using the `union()` method from `lubridate`:

```{r lubridate-union}
union(greece, brazil)
```

The resulting interval includes the intervening time between the disjoint vacations. The `phinterval` package provides a drop-in replacement, `phint_union()`, which accepts the same arguments but returns a *phinterval* vector.

```{r phinterval-union}
phint_union(greece, brazil)
```

The result is a disjoint time span, preserving the gap between the two vacations.

In simple calculations, this distinction can easily lead to unexpected results. For example, to calculate the number of days that either employee is out of the office, one might take the union of their vacation spans and then calculate the duration in days: 

```{r duration}
as_duration(union(greece, brazil)) / ddays()
as_duration(phint_union(greece, brazil)) / ddays()
```

While experienced `lubridate` users can anticipate and work around these cases, `phinterval` provides an intuitive alternative which can easily be substituted into existing analyses.

# References
