---
title: 'phinterval: An R package for representing and manipulating time spans with gaps'
tags:
  - R
  - dates
  - timespans
  - set operations
authors:
  - name: Ethan Sansom
    orcid: 0009-0000-1573-0186
    affiliation: 1
affiliations:
  - name: "Department of Statistical Sciences, University of Toronto"
    index: 1
date: "2025-12-18"
bibliography: paper.bib
output:
  # rticles::joss_article:
  #  keep_tex: false
  md_document:
     preserve_yaml: TRUE
     variant: "markdown_strict"
journal: JOSS
---

# Summary

`phinterval` is an R [@R] package for representing and manipulating time spans 
that may include gaps. It implements the *phinterval* vector class, designed as 
a generalization of the `lubridate` [@lubridate] package's *Interval* class.

While existing interval classes represent contiguous spans of time defined
by a start and an end point, each element of a *phinterval* is a union of zero
or more disjoint intervals. This allows empty time spans (e.g., the
intersection of two non-overlapping events) and disjoint time spans (e.g., a
student’s periods of enrollment in school separated by breaks) to be treated
as first-class objects and manipulated using well-defined set operations.

Functionality for working with these time spans includes:

- Set Operations: Vectorized union, intersection, difference, and complement.
- Merging of overlapping and adjacent intervals into minimal non-overlapping sets.
- Tests for whether time spans, dates, or times fall within one another or overlap.

The package is designed to work seamlessly with `lubridate`: all `phinterval`
functions accept either *Interval* or *phinterval* vectors as input, enabling
analysts to safely drop `phinterval` functions into their existing workflows.

# Statement of Need

Accurately representing and manipulating dates and times is challenging due to
complexities such as time zone adjustments, daylight saving transitions, and leap 
years or seconds. As a result, temporal data analysis is a common source of 
frustration and error-prone or difficult-to-maintain code [@lubridate; @itsabouttime]. 
Several R packages provide intuitive interfaces that that handle these complexities 
automatically. In particular, `lubridate`, which is widely used for date-time 
manipulation in R, and `ivs` [@ivs] simplify the representation and manipulation 
of time spans, reducing users' cognitive load and the likelihood of mistakes.

To the author's knowledge, however, no existing package provides a first-class
representation of empty or disjoint time spans. Users encountering these spans,
for example, the intersection of two non-overlapping intervals, receive an error 
(as in `ivs`) or unintuitive results (as in `lubridate`), forcing workarounds and
potentially leading to uncaught mistakes. 

The `phinterval` package addresses this gap by providing explicit representations 
of disjoint and empty time spans and defining set operations whose results are 
always valid *phinterval* objects. By closely mirroring the `lubridate` interface 
and accepting *Interval* vectors as input, `phinterval` minimizes adoption costs 
while enabling safe and intuitive manipulation of arbitrary time spans. The package 
is intended for analysts and researchers working with temporal data in settings
such as event studies and observational data analysis.

# Examples

The following examples demonstrate how `phinterval` functions can be used as 
drop-in replacements for `lubridate` operations, providing correct results for 
empty or disjoint intervals that would otherwise produce errors or ambiguous 
output.

```{r, include = FALSE}
# TODO: Temporary fix, remove this later
if (!rlang::is_installed("phinterval")) {
  devtools::install_github("EthanSansom/phinterval")
}
```

```{r dependencies, warning = FALSE, message = FALSE}
library(phinterval)

jan_1_to_9 <- interval(as.Date("2000-01-01"), as.Date("2000-01-09"))
jan_2_to_3 <- interval(as.Date("2000-01-02"), as.Date("2000-01-03"))
jan_5_to_9 <- interval(as.Date("2000-01-05"), as.Date("2000-01-09"))
```

In `lubridate`, the intersection of non-overlapping intervals returns an object 
with missing endpoints, resulting in ambiguity between genuinely empty time
spans and missing data. `phinterval` explicitly represents empty time spans
as a `<hole>`.

```{r intersect}
lubridate::intersect(jan_2_to_3, jan_5_to_9)
phint_intersect(jan_2_to_3, jan_5_to_9)
```

Standard interval classes cannot represent gaps within a single observation, 
but `phinterval` handles these naturally.

```{r setdiff}
try(lubridate::setdiff(jan_1_to_9, jan_2_to_3))
phint_setdiff(jan_1_to_9, jan_2_to_3)
```

In addition to standard set operations, `phinterval` provides specialized functions 
for working with intervals with gaps. For example, `phint_squash()` flattens a 
vector of intervals into a minimal set of non-overlapping time spans, while 
`phint_invert()` returns the gaps between disjoint intervals.

These functions are useful for summarizing longitudinal event data, such as 
periods of employment and intervening unemployment:

```{r}
jobs <- dplyr::tribble(
  ~name,   ~job_title,             ~start,        ~end,
  "Greg",  "Mascot",               "2018-01-01",  "2018-06-03",
  "Greg",  "Chief of Staff",       "2019-03-01",  "2020-11-28",
  "Tom",   "Chairman",             "2019-05-01",  "2020-11-10",
  "Tom",   "CEO",                  "2020-11-10",  "2020-12-31",
  "Shiv",  "Political Consultant", "2017-01-01",  "2019-04-01"
)

employment <- jobs |>
  dplyr::mutate(span = interval(start, end)) |>
  dplyr::group_by(name) |>
  dplyr::summarize(employment = phint_squash(span))

employment
```

All `phinterval` functions are vectorized, making them well suited for data
analysis.

```{r}
employment |>
  dplyr::mutate(
    # How long were workers employed?
    n_emp_days = employment / lubridate::ddays(1),

    # Where were the gaps in workers' employment?
    unemployment = phint_invert(employment)
  ) |>
  dplyr::select(name, n_emp_days, unemployment)
```

# References
